---
title: "BRCA-Project"
output: html_notebook
---


# PROJECT 1 — TCGA-BRCA \| Breast Invasive Carcinoma

This project analyses RNA-seq gene expression data from The Cancer Genome Atlas
(TCGA) BRCA cohort. The goal is to predict PAM50 molecular subtypes and
integrate clinical metadata for biological interpretation.

**Reference:** The Cancer Genome Atlas Network (2012). Comprehensive molecular
portraits of human breast tumours. *Nature*, 490, 61–70.

---

## Setup — Load Libraries & Data

We load all required packages and the BRCA dataset provided
as an `.RData` file.
## Cleaning annotations

```{r}
load("BRCA_data.RData")
library(dplyr) 
library(ggplot2) 
library(pheatmap) 
library(caret)
library(tibble)
library(tidyr)
library(cluster)
```

=======

# Part I - Gene Expression Analysis 
## Step 1 — Data Structure Investigation
### How many samples and genes are in the expression matrix?

```{r}
# The dim() function returns the dimensions of your dataset: [1] Number of Rows (Genes), [2] Number of Columns. (Samples)

dim <- dim(brca_exp_clean)
df_dims <- data.frame("Genes" = dim[1], "Samples" = dim[2]) ; df_dims
```
### How many annotation columns are retained after cleaning to paper_ columns?
We keep only the curated publication-level columns (those starting with
`paper_`) plus the `barcode` identifier. This filters out internal TCGA
processing columns that are not relevant for our analysis.
```{r}
# --- 1. Samples and Genes in the Expression Matrix ---

# indexes the annotations for the columns starting with paper, adding a barcode column for counting.
anot_paper <- brca_anot_clean[, c("barcode", grep("^paper_", names(brca_anot_clean), value = TRUE))]

# check how many columns start with paper_
num_paper_cols <- ncol(anot_paper) - 1 # Subtract 1 to exclude the 'barcode' column
cat("Number of paper_ columns retained:", num_paper_cols, "\n")

dim <- dim(brca_anot_clean)
df_dims <- data.frame("Samples" = dim[1], "annotations" = dim[2]) ; df_dims

```

### What is the sample count per PAM50 subtype (paper_BRCA_Subtype_PAM50)?

papter_BRCA-Subtype_PAM50 is our primary target variable. The PAM50 classifier assigns each tumour to one of five intrinsic subtypes: Basal, Her2, LumA, LumB, and Normal-like.


```{r}
# Shows all PAM50 subtypes in a table. The "ifany" argument adds a column with the NA count if there are any NAs available.

subtype_counts <- table(anot_paper$paper_BRCA_Subtype_PAM50, useNA = "ifany")

print("Sample count per PAM50 subtype:")
print(subtype_counts)
```

```{r}
## Finding the Valid Patients
# The primary target for BRCA is paper_BRCA_Subtype_PAM50
valid_target_samples <- sum(!is.na(anot_paper$paper_BRCA_Subtype_PAM50))
cat("Samples with non-NA PAM50 subtype:", valid_target_samples, "\n")

# Alternative
data.frame("Valid_PAM50_subtypes" = valid_target_samples, "NA" = length(anot_paper$paper_BRCA_Subtype_PAM50)-valid_target_samples)

```

## significant breast cancer genes from the paper:

PIK3CA, PTEN, AKT1, TP53, GATA3, CDH1, RB1, MLL3, MAP3K1, CDKN1B, TBX3, RUNX1, CBFB, AFF2, PIK3R1, PTPN22, PTPRD, NF1, SF3B1 and CCND3


## How many annotation columns are retained after cleaning to paper\_ columns?

```{r}

#---------------------Not Needed-------------------


# check how many columns start with paper_
num_paper_cols <- ncol(anot_paper) - 1 # Subtract 1 to exclude the 'barcode' column
cat("Number of paper_ columns retained:", num_paper_cols, "\n")
```


## Step 2 — Expression Distribution Analysis

### Randomly select 10 genes and visualize their expression distributions across all samples.

```{r}

set.seed(161) # For reproducibility
random_exp <- brca_exp_clean %>% slice_sample(n=10) # Subsetting the matrix to keep randomly selected genes
random_exp_long <- random_exp %>%
  pivot_longer(cols = where(is.numeric), 
               names_to = "Sample", 
               values_to = "Expression") # Reshaping data for ggplot2

ggplot(random_exp_long, aes(x = gene, y = Expression, fill = gene)) +
geom_boxplot(alpha = 0.7, outlier.size = 0.5) +  
  theme_minimal() +                                
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"), 
        legend.position = "none") +                
  labs(title = "Expression Distribution of 10 Randomly Selected Genes",
       x = "Gene",
       y = "Normalized Expression Level")


```


### Select the top 5 and bottom 5 most variable genes (by standard deviation); plot their distributions and compare.

```{r}
# Calculate gene variability (standard deviation)
# Calculate standard deviation for each gene (across all samples)
# Exclude the "gene" column, only use expression columns
gene_variability <- brca_exp_clean %>%
  mutate(gene_sd = apply(select(., -gene), 1, sd, na.rm = TRUE)) %>%
  select(gene, gene_sd) %>%
  arrange(desc(gene_sd))

# Top 5 most variable
top5_variable <- head(gene_variability, 5)
cat("Top 5 Most Variable Genes:\n")
print(top5_variable)

# Bottom 5 least variable
bottom5_variable <- tail(gene_variability, 5)
cat("\nBottom 5 Least Variable Genes:\n")
print(bottom5_variable)

# Combine gene lists
variable_genes <- c(top5_variable$gene, bottom5_variable$gene)

# Extract and transform data
var_exp <- brca_exp_clean %>%
  filter(gene %in% variable_genes) %>%
  pivot_longer(-gene, names_to = "Sample", values_to = "Expression") %>%
  mutate(Variability = ifelse(gene %in% top5_variable$gene, "High", "Low"))

# Plot comparison
ggplot(var_exp, aes(x = reorder(gene, Expression, FUN = median), 
                    y = Expression, 
                    fill = Variability)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_manual(values = c("High" = "#E74C3C", "Low" = "#3498DB")) +
  labs(title = "Expression Distribution: High vs Low Variability Genes",
       subtitle = "Top 5 and Bottom 5 genes by standard deviation",
       x = "Gene",
       y = "Normalized Expression (log2)",
       fill = "Variability") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## Step 3 — Correlation Analysis
## Prerequisite — Sample Alignment

Before running any analysis that combines the expression matrix with the annotation file, we must ensure that both objects contain **exactly the same samples in the same order**. Mismatches would silently corrupt all downstream results.
```{r}
#First set the column "gene" as rownames 
brca_exp_clean_ctr <- brca_exp_clean %>%
  column_to_rownames("gene")
```

```{r} 
# TRUE = Reihenfolge stimmt bereits überein, kein Alignment nötig
all(colnames(brca_exp_clean_ctr) == anot_paper$barcode)
``` 
Because the the result says "TRUE" we don't need to correct the alginment. The order and dimensions are aligned. Else we would have to write the following aligmentprompt: 

```{r}
#Find samples that appear in BOTH the expression matrix "brca_exp_clean" AND the significant annotation file "anot_paper"
common_samples <- intersect(colnames(brca_exp_clean_ctr), anot_paper$barcode)

#Subset both objects to the shared samples (same order in both)
expr_clean <- brca_exp_clean_ctr[ , common_samples]
anot_clean <- anot_paper[match(common_samples, anot_paper$barcode), ]
```

```{r}
#testing if samples have been removed by comparing dimensions before and after 
# before
dim(brca_exp_clean_ctr)
dim(anot_paper)

# after 
dim(expr_clean)
dim(anot_clean)

#this again demonstrates that the last step was unnecessary
```

### Compute pairwise Pearson correlations across all samples using the full gene set.
We compute **pairwise Pearson correlations** between all samples using the full gene set. The resulting n×n correlation matrix is visualised as a heatmap. Samples are annotated by PAM50 subtype so we can judge whether the correlation structure reflects known biology.

The Pearson correlation measures linear co-expression: two samples with similar expression profiles across all genes will have a correlation close to 1. If PAM50 subtypes are transcriptionally distinct, we expect to see high within-subtype correlations forming visible blocks along the diagonal.

```{r}
#calculate Pearson-correlation matrix between all samples. 
#convert to numeric matrix (required by c())
expr_mat <- as.matrix(brca_exp_clean_ctr)

#cor() applied column-wise computes correlation betweent samples. Each column = one sample's expression profile across all genes
cor_mat <- cor(expr_mat, method = "pearson")
```

```{r}
#unwichtig -> gibt mir die dimensions der correlation matrix was logischerweise 1099x1099 ist da das die sample zahl ist
# gibt auch die value range
cat("Correlation matrix dimensions:", dim(cor_mat)[1], "x", dim(cor_mat)[2], "\n")
cat("Value range: [", round(min(cor_mat), 3), ",", round(max(cor_mat), 3), "]\n")
```
### Visualize as a correlation heatmap. What structure do you observe? Are there sub- blocks?
```{r}
#colors for PAM50-Subtpyes
pam50_colors <- c(
  Basal = "red",
  Her2 = "orange",
  LumA = "blue",
  LumB = "green",
  Normal = "purple"
)

# annotation for heatmap
sample_annotation <- data.frame(
  PAM50 = anot_clean$paper_BRCA_Subtype_PAM50,
  row.names = anot_clean$barcode
  )

# plot heatmap
pheatmap(
  cor_mat,
  annotation_col = sample_annotation,
  annotation_row = sample_annotation,
  annotation_colors = list(PAM50 = pam50_colors),
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("white","orange","darkred"))(100),
  main = "Pairwise Pearson Correlation",
  treeheight_row = 30,
  treeheight_col = 30
)
```



```{r}
# Arijaan block
gene_data <- data[, gene_columns] 

# 2. Transpose the matrix (patients become columns, genes become rows)
transposed_genes <- t(gene_data)

# 3. Compute the 1099 x 1099 patient-to-patient correlation matrix
patient_cor_matrix <- cor(transposed_genes, method = "pearson")


corrs <- cor(brca_exp_clean[,-1],method="pearson")
```


### Interpret: does the correlation structure correspond to PAM50 subtypes?

```{r}

pheatmap(corrs, cluster_rows = FALSE,
  cluster_cols = FALSE, show_rownames = F,
  show_colnames = F)

# Jungs ich weiß nicht ob das hier richtig ist, aber so siehts gerade mit dem heatmap aus: -@Arijaan

```


## Step 4 — Hierarchical Clustering

```{r}
<<<<<<< HEAD
#sample_dist <- dist(t(brca_exp_clean), method = "euclidean")

#Perform hierarchical clustering

#hc <- hclust(sample_dist, method = "ward.D2")
#k <- 5
#ect.hc(hc, k = k, border = "red")
#hc_clusters <- cutree(hc, k = k)
#print(table(hc_clusters))
=======
# Check the dimensions
dim(exp_t)

# Check for NAs
sum(is.na(exp_t))

# Check for infinite values
sum(is.infinite(exp_t))

# Check data types of each column
str(exp_t[, 1:5])  # look at first 5 columns

# Check if any columns are non-numeric
any(!apply(exp_t, 2, is.numeric))

```
```{r}
# If brca_exp_clean is a data frame with a gene name column
brca_exp_clean_fixed <- brca_exp_clean %>%
  column_to_rownames("gene") %>%   # move gene names to rownames
  mutate(across(everything(), as.numeric))  # ensure all values are numeric

# Verify
str(brca_exp_clean_fixed[1:3, 1:3])  # should say num
dim(brca_exp_clean_fixed)            # should be 1099 x 2994 (one less row now)

# Now safely proceed

```


### Cluster samples using hclust.
```{r step4-hierarchical-clustering, fig.width=10, fig.height=6}

# chunk 2 - evaluate k and cut dendrogram
library(cluster)

exp_t <- t(brca_exp_clean_fixed)
D <- dist(exp_t, method = "euclidean")
hc <- hclust(D, method = "ward.D2")

# evaluate k using silhouette
km <- kmeans(x = exp_t, centers = 3, nstart = 10)
s <- silhouette(km$cluster, D)
plot(s, col = 'blue')
>>>>>>> 6b4c5be (part 4 continuation)


```


### Cut the dendrogram to obtain k clusters (choose k; justify your choice).

```{r}
# cut the dendrogram
k_chosen <- 3
hc_clusters <- cutree(hc, k = k_chosen)

# visualize
plot(hc,
     labels = FALSE,
     main = "Hierarchical clustering of BRCA samples",
     xlab = "Samples",
     ylab = "Height")

# add rectangles showing the k clusters on the dendrogram
rect.hclust(hc, k = k_chosen, border = "red")

# check cluster sizes
table(hc_clusters)
```


### Compare cluster membership with PAM50 subtypes.

## Step 5 — PCA Visualization

### Run prcomp on the transposed expression matrix (samples as rows).

### Plot PC1 vs. PC2 and PC1 vs. PC3, colored by PAM50 subtype and by hierarchical cluster.

### Report the variance explained by the first three PCs. What distinguishes the subtypes?

## Step 6 — Differential Expression

### Use a Wilcoxon rank-sum test (one-vs-rest) to identify marker genes for each PAM50 subtype.

```{r}
wilcox.test()
```


### Apply FDR correction (BH). Report the top 20 genes per subtype by adjusted p-value and log2 fold-change.

### Generate an annotated heatmap (rows = top DEGs, columns = samples, annotation bars = PAM50 subtype and pathologic stage).

## Step 7 — Prediction Models

### Split data 70/30 (stratified by subtype).

```{r}

# Function to remove columns with lots of NAs (more than half of the entries).
# Used later on in the script.
#nacol_rm <- function(df, p){
#  names <- c()
#  valid_colnames <- sapply(1:ncol(df), function(i){
#  ifelse(sum(is.na(df[,i])) > round(nrow(df)*p), NA,
#         names[i]<-colnames(df[,i])
#    )
#   }
#  )

#valid_colnames <- na.omit(valid_colnames)
#df[,valid_colnames]
#}

# ---------alternative to the function: (delete if the function works ---LOL

nacol_rm <- function(df,p){
too_many_nas <- colSums(is.na(df)) > round(nrow(df)*p)
bad_column_names <- colnames(df)[too_many_nas]
df <- df[, !too_many_nas]
}
```

For binary analysis of the data, we need to annotate the subtypes properly, so that we have either a "normal" or a "tumor". For this, we can use `dplyr`:

```{r}

# Coverstion of subtypes into binary:
normal <- brca_anot_clean %>% filter(paper_BRCA_Subtype_PAM50 == "Normal") 
tumor <- brca_anot_clean %>% filter(!paper_BRCA_Subtype_PAM50 == "Normal") 
tumor$paper_BRCA_Subtype_PAM50 <- "tumor"
binary_data <- rbind(normal,tumor)

# Rownames to cols
 
binary_data <- binary_data %>% 
  column_to_rownames("barcode")

binary_data[numeric_cols]
# Convert to numeric
#numeric_cols <- setdiff(names(binary_data), "paper_BRCA_Subtype_PAM5")

#binary_data[numeric_cols] <- lapply(binary_data[,numeric_cols], function(x) as.numeric(as.character(x)))


## AI version:

binary_data <- binary_data %>%
  # Convert everything EXCEPT the target column to numeric
  mutate(across(-paper_BRCA_Subtype_PAM50, ~ as.numeric(as.character(.))))

```


```{r}

# Using caret to create the test and training data
trainIndex <- createDataPartition(binary_data$paper_BRCA_Subtype_PAM50, p = 0.7, list = FALSE)


train_data <- binary_data[trainIndex, ]
test_data  <- binary_data[-trainIndex, ]

# Clearing out columns with lots of NAs 
train_data <- nacol_rm(df=train_data, p=0.5) # proportion of cols with NAs
test_data <- nacol_rm(df=test_data, p=0.5)

```


```{r}

##################### Gemini glm #############################

# 1. Conversion of subtypes into binary
normal <- brca_anot_clean %>% filter(paper_BRCA_Subtype_PAM50 == "Normal") 
tumor <- brca_anot_clean %>% filter(paper_BRCA_Subtype_PAM50 != "Normal") # Safer than !col == "Normal"
tumor$paper_BRCA_Subtype_PAM50 <- "tumor"
binary_data <- rbind(normal, tumor)

# 2. Rownames to cols
binary_data <- binary_data %>% 
  column_to_rownames("barcode")

# 3. Handle NAs and Factors safely at the dataframe level
binary_data <- binary_data %>% 
  filter(!is.na(paper_BRCA_Subtype_PAM50)) %>% 
  mutate(paper_BRCA_Subtype_PAM50 = factor(paper_BRCA_Subtype_PAM50, levels = c("Normal", "tumor")))

# 4. Get numeric columns (Typo fixed)
numeric_cols <- setdiff(names(binary_data), "paper_BRCA_Subtype_PAM50")

# (Optional: Convert to numeric if you uncommented your lapply)
# binary_data[numeric_cols] <- lapply(binary_data[numeric_cols], function(x) as.numeric(as.character(x)))

# 5. Split Data
set.seed(123) # Always set a seed for reproducibility
trainIndex <- createDataPartition(binary_data$paper_BRCA_Subtype_PAM50, p = 0.7, list = FALSE)

train_data <- binary_data[trainIndex, ]
test_data  <- binary_data[-trainIndex, ]

# 6. Correct NA handling (Preventing Data Leakage)
# Calculate which columns to drop based ONLY on training data
too_many_nas <- colSums(is.na(train_data)) > round(nrow(train_data) * 0.5)
cols_to_drop <- colnames(train_data)[too_many_nas]

# Drop the exact same columns from BOTH datasets
train_data <- train_data[, !colnames(train_data) %in% cols_to_drop]
test_data  <- test_data[, !colnames(test_data) %in% cols_to_drop]

# 7. Model
model_logit <- glm(paper_BRCA_Subtype_PAM50 ~ ., data = train_data, family = "binomial")

summary(model_logit)
```

### Binary task — tumor vs. normal: Logistic Regression, Random Forest, k-NN. Report confusion matrices and AUC-ROC.

#### Logistic regression

```{r Logistic regression}

# Creating the model using the glm function from the stats library.

# NAs can be inputted here since the function takes out all NAs. 
model_logit <- glm(paper_BRCA_Subtype_PAM50 ~ ., data = train_data, family = "binomial")

summary(model_logit)

# Predict
prob_logit <- predict(model_logit, test, type = "response")
pred_logit <- ifelse(prob_logit > 0.5, "normal", "tumor")
pred_logit <- factor(pred_logit, levels = c("normal", "tumor"))

confusionMatrix(pred_logit$paper_vital_status, test_data$paper_vital_status)

```

#### K-Nearest Neighbor:
```{r}
# Train k-NN
set.seed(123)
model_knn <- train(
  Class ~ .,
  data = train_norm,
  method = "knn",
  tuneGrid = expand.grid(k = 1:10),
  trControl = trainControl(method = "cv", number = 5, classProbs = TRUE)
)
```

```{r}
# Predict
knn_probs <- predict(model_knn, test_norm, type = "prob")[, "malignant"]
pred_knn <- ifelse(knn_probs > 0.5, "malignant", "benign")
pred_knn <- factor(pred_knn, levels = c("malignant", "benign"))
confusionMatrix(pred_knn, test_norm$Class)

```
```{r}
# Normalize features (excluding Class)
preproc <- preProcess(train_data[, -10], method = c("center", "scale"))
train_norm <- predict(preproc, train_data[, -10])
test_norm <- predict(preproc, test_data[, -10])

# Add Subtype back
train_norm <- train_data
test_norm <- test_data
```


```{r}
# Plot k selection
ggplot(model_knn$results, aes(x = k, y = Accuracy)) +
  geom_line(col="red3") +
  geom_point() +
  labs(title = "k-NN Accuracy vs. Number of Neighbors")
```

#### Random forest


-----------------------------------------------------------------



### Multiclass task — predict PAM50 subtype: Random Forest, Multinomial Logistic Regression, k-NN.

#### Random Forest:

```{r}
library(randomForest)
# appending meaningfull expression data of the top 100 genes to the annotations dataset for increasing accuarcy.

sig_genes <- head(gene_variability, 100)
sig_genes <- column_to_rownames(sig_genes, "gene")
dat <- brca_anot_clean %>% 
  column_to_rownames("barcode") %>% 
  filter(!is.na(paper_BRCA_Subtype_PAM50)) %>% 
  mutate(exp = t(sig_genes))
# Fit the random forest model
# Using caret to create the test and training data

trainIndex <- createDataPartition(dat$paper_BRCA_Subtype_PAM50, p = 0.7, list = FALSE)


train_data <- binary_data[trainIndex, ]
test_data  <- binary_data[-trainIndex, ]

# Clearing out columns with lots of NAs 
train_data <- nacol_rm(df=train_data, p=0.5) # proportion of cols with NAs
test_data <- nacol_rm(df=test_data, p=0.5)

model_rf <- randomForest(paper_BRCA_Subtype_PAM50 ~ ., data = train_data, importance = TRUE, ntree = 500)

# Feature importance plot
varImpPlot(model_rf, main = "Feature Importance")

# Predict
pred_rf <- predict(model_rf, test_data)
confusionMatrix(pred_rf, test_data$paper_br) # enter subtype 

confusionMatrix(pred_rf, test_data) # enter subtype
```

### For each model, report: accuracy, F1, per-class precision/recall.

## Step 8 — Method Comparison

### Compare model performance systematically in a summary table.

### Evaluate whether the top DEGs (Step 6) rank highly as predictive features (Step 7 feature importance). Overlap analysis with UpSet plot or Venn diagram.

# Part II — Clinical Data Analysis

## Step A — Annotation Cleaning

### Tidy the annotation file to keep only curated, paper-level columns.

### anot_clean \<- brca_anot_clean[, c("barcode", grep("\^paper\_", names(brca_anot_clean), value = TRUE))]

### How many paper\_ columns are retained? What categories of information do they cover?

### Are there samples with missing subtype assignments? How should these be handled?

## Step B — Clinical Variable Exploration

### Explore the clinical metadata as a standalone dataset before merging with expression data.

### Explore paper_pathologic_stage, paper_BRCA_Pathology (IDC vs ILC), paper_age_at_initial_pathologic_diagnosis.

### Tabulate paper_vital_status and compute median follow-up time.

### Examine multi-omic cluster columns (paper_CNV.Clusters, paper_mRNA.Clusters, paper_Protein.Clusters) — how do they relate to PAM50?

## Step C — Clinical Feature Classification

### Predict a binary or multi-class clinical outcome using clinical variables only (no expression data).

### Define the outcome variable and apply appropriate encoding.

### Select a small set of meaningful features (age, stage, grade, subtype, etc.) and handle missing values by imputation or exclusion.

### Train/test split (70/30 or 5-fold CV). Fit: Logistic Regression, Random Forest, k-NN.

### Report confusion matrices, accuracy, sensitivity/specificity, and AUC-ROC for each model.

### Identify the most predictive clinical features: variable importance from RF, and coefficients from Logistic Regression. Note that k-NN does not provide native feature importance.

## Step D — Integrative Analysis (Expression + Clinical)

### Correlate top DEGs from Part I with continuous clinical variables (e.g., age) using Spearman's rho.

### Test whether top DEGs differ significantly across clinical stages.

### Build a multimodal classifier: combine top expression features (from Step 7 / RF importance) with key clinical features. Does adding clinical data improve performance?

### Visualize: boxplots of top expression features stratified by clinical subgroups; annotated heatmap with both molecular and clinical annotation bars.

### 

## Step E - Discussion & Interpretation

### Critical thinking questions

### Are the molecular subtypes (from expression clustering) concordant with clinical

staging? Where do they disagree?

### Which data modality — expression or clinical — gives better predictive accuracy for

your outcome variable?

### What biological mechanisms might explain the top DEGs found between clinical

subgroups?

### What are the limitations of the TCGA cohort for clinical prediction (sample size per

stage, selection of academic-centre patients, etc.)?
