---
title: "BRCA-Project"
output: html_notebook
---


# PROJECT 1 — TCGA-BRCA \| Breast Invasive Carcinoma

This project analyses RNA-seq gene expression data from The Cancer Genome Atlas
(TCGA) BRCA cohort. The goal is to predict PAM50 molecular subtypes and
integrate clinical metadata for biological interpretation.

**Reference:** The Cancer Genome Atlas Network (2012). Comprehensive molecular
portraits of human breast tumours. *Nature*, 490, 61–70.

---

## Setup — Load Libraries & Data

We load all required packages and the BRCA dataset provided
as an `.RData` file.
## Cleaning annotations

```{r}
load("BRCA_data.RData")
install.packages("pROC")
library(dplyr) 
library(ggplot2) 
library(pheatmap) 
library(caret)
library(tibble)
library(tidyr)
library(cluster)
library(pROC)
```

=======

# Part I - Gene Expression Analysis 
## Step 1 — Data Structure Investigation
### How many samples and genes are in the expression matrix?

```{r}
# The dim() function returns the dimensions of your dataset: [1] Number of Rows (Genes), [2] Number of Columns. (Samples)

dim <- dim(brca_exp_clean)
df_dims <- data.frame("Genes" = dim[1], "Samples" = dim[2]) ; df_dims
```
### How many annotation columns are retained after cleaning to paper_ columns?
We keep only the curated publication-level columns (those starting with
`paper_`) plus the `barcode` identifier. This filters out internal TCGA
processing columns that are not relevant for our analysis.
```{r}
# --- 1. Samples and Genes in the Expression Matrix ---

# indexes the annotations for the columns starting with paper, adding a barcode column for counting.
anot_paper <- brca_anot_clean[, c("barcode", grep("^paper_", names(brca_anot_clean), value = TRUE))]

# check how many columns start with paper_
num_paper_cols <- ncol(anot_paper) - 1 # Subtract 1 to exclude the 'barcode' column
cat("Number of paper_ columns retained:", num_paper_cols, "\n")

dim <- dim(brca_anot_clean)
df_dims <- data.frame("Samples" = dim[1], "annotations" = dim[2]) ; df_dims

```

### What is the sample count per PAM50 subtype (paper_BRCA_Subtype_PAM50)?

papter_BRCA-Subtype_PAM50 is our primary target variable. The PAM50 classifier assigns each tumour to one of five intrinsic subtypes: Basal, Her2, LumA, LumB, and Normal-like.


```{r}
# Shows all PAM50 subtypes in a table. The "ifany" argument adds a column with the NA count if there are any NAs available.

subtype_counts <- table(anot_paper$paper_BRCA_Subtype_PAM50, useNA = "ifany")

print("Sample count per PAM50 subtype:")
print(subtype_counts)
```

```{r}
## Finding the Valid Patients
# The primary target for BRCA is paper_BRCA_Subtype_PAM50
valid_target_samples <- sum(!is.na(anot_paper$paper_BRCA_Subtype_PAM50))
cat("Samples with non-NA PAM50 subtype:", valid_target_samples, "\n")

# Alternative
data.frame("Valid_PAM50_subtypes" = valid_target_samples, "NA" = length(anot_paper$paper_BRCA_Subtype_PAM50)-valid_target_samples)

```

## significant breast cancer genes from the paper:

PIK3CA, PTEN, AKT1, TP53, GATA3, CDH1, RB1, MLL3, MAP3K1, CDKN1B, TBX3, RUNX1, CBFB, AFF2, PIK3R1, PTPN22, PTPRD, NF1, SF3B1 and CCND3


## How many annotation columns are retained after cleaning to paper\_ columns?

```{r}

#---------------------Not Needed-------------------


# check how many columns start with paper_
num_paper_cols <- ncol(anot_paper) - 1 # Subtract 1 to exclude the 'barcode' column
cat("Number of paper_ columns retained:", num_paper_cols, "\n")
```


## Step 2 — Expression Distribution Analysis

### Randomly select 10 genes and visualize their expression distributions across all samples.

```{r}

set.seed(161) # For reproducibility
random_exp <- brca_exp_clean %>% slice_sample(n=10) # Subsetting the matrix to keep randomly selected genes
random_exp_long <- random_exp %>%
  pivot_longer(cols = where(is.numeric), 
               names_to = "Sample", 
               values_to = "Expression") # Reshaping data for ggplot2

ggplot(random_exp_long, aes(x = gene, y = Expression, fill = gene)) +
geom_boxplot(alpha = 0.7, outlier.size = 0.5) +  
  theme_minimal() +                                
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"), 
        legend.position = "none") +                
  labs(title = "Expression Distribution of 10 Randomly Selected Genes",
       x = "Gene",
       y = "Normalized Expression Level")


```


### Select the top 5 and bottom 5 most variable genes (by standard deviation); plot their distributions and compare.

```{r}
# Calculate gene variability (standard deviation)
# Calculate standard deviation for each gene (across all samples)
# Exclude the "gene" column, only use expression columns
gene_variability <- brca_exp_clean %>%
  mutate(gene_sd = apply(select(., -gene), 1, sd, na.rm = TRUE)) %>%
  select(gene, gene_sd) %>%
  arrange(desc(gene_sd))

# Top 5 most variable
top5_variable <- head(gene_variability, 5)
cat("Top 5 Most Variable Genes:\n")
print(top5_variable)

# Bottom 5 least variable
bottom5_variable <- tail(gene_variability, 5)
cat("\nBottom 5 Least Variable Genes:\n")
print(bottom5_variable)

# Combine gene lists
variable_genes <- c(top5_variable$gene, bottom5_variable$gene)

# Extract and transform data
var_exp <- brca_exp_clean %>%
  filter(gene %in% variable_genes) %>%
  pivot_longer(-gene, names_to = "Sample", values_to = "Expression") %>%
  mutate(Variability = ifelse(gene %in% top5_variable$gene, "High", "Low"))

# Plot comparison
ggplot(var_exp, aes(x = reorder(gene, Expression, FUN = median), 
                    y = Expression, 
                    fill = Variability)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_manual(values = c("High" = "#E74C3C", "Low" = "#3498DB")) +
  labs(title = "Expression Distribution: High vs Low Variability Genes",
       subtitle = "Top 5 and Bottom 5 genes by standard deviation",
       x = "Gene",
       y = "Normalized Expression (log2)",
       fill = "Variability") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## Step 3 — Correlation Analysis
## Prerequisite — Sample Alignment

Before running any analysis that combines the expression matrix with the annotation file, we must ensure that both objects contain **exactly the same samples in the same order**. Mismatches would silently corrupt all downstream results.
```{r}
#First set the column "gene" as rownames 
brca_exp_clean_ctr <- brca_exp_clean %>%
  column_to_rownames("gene")
```

```{r} 
# TRUE = Reihenfolge stimmt bereits überein, kein Alignment nötig
all(colnames(brca_exp_clean_ctr) == anot_paper$barcode)
``` 
Because the the result says "TRUE" we don't need to correct the alginment. The order and dimensions are aligned. Else we would have to write the following aligmentprompt: 

```{r}
#Find samples that appear in BOTH the expression matrix "brca_exp_clean" AND the significant annotation file "anot_paper"
common_samples <- intersect(colnames(brca_exp_clean_ctr), anot_paper$barcode)

#Subset both objects to the shared samples (same order in both)
expr_clean <- brca_exp_clean_ctr[ , common_samples]
anot_clean <- anot_paper[match(common_samples, anot_paper$barcode), ]
```

```{r}
#testing if samples have been removed by comparing dimensions before and after 
# before
dim(brca_exp_clean_ctr)
dim(anot_paper)

# after 
dim(expr_clean)
dim(anot_clean)

#this again demonstrates that the last step was unnecessary
```

### Compute pairwise Pearson correlations across all samples using the full gene set.
We compute **pairwise Pearson correlations** between all samples using the full gene set. The resulting n×n correlation matrix is visualised as a heatmap. Samples are annotated by PAM50 subtype so we can judge whether the correlation structure reflects known biology.

The Pearson correlation measures linear co-expression: two samples with similar expression profiles across all genes will have a correlation close to 1. If PAM50 subtypes are transcriptionally distinct, we expect to see high within-subtype correlations forming visible blocks along the diagonal.

```{r}
#calculate Pearson-correlation matrix between all samples. 
#convert to numeric matrix (required by c())
expr_mat <- as.matrix(brca_exp_clean_ctr)

#cor() applied column-wise computes correlation betweent samples. Each column = one sample's expression profile across all genes
cor_mat <- cor(expr_mat, method = "pearson")
```

```{r}
#unwichtig -> gibt mir die dimensions der correlation matrix was logischerweise 1099x1099 ist da das die sample zahl ist
# gibt auch die value range
cat("Correlation matrix dimensions:", dim(cor_mat)[1], "x", dim(cor_mat)[2], "\n")
cat("Value range: [", round(min(cor_mat), 3), ",", round(max(cor_mat), 3), "]\n")
```
### Visualize as a correlation heatmap. What structure do you observe? Are there sub- blocks?
```{r}
#colors for PAM50-Subtpyes
pam50_colors <- c(
  Basal = "red",
  Her2 = "orange",
  LumA = "blue",
  LumB = "green",
  Normal = "purple"
)

# annotation for heatmap
sample_annotation <- data.frame(
  PAM50 = anot_clean$paper_BRCA_Subtype_PAM50,
  row.names = anot_clean$barcode
  )

# plot heatmap
pheatmap(
  cor_mat,
  annotation_col = sample_annotation,
  annotation_row = sample_annotation,
  annotation_colors = list(PAM50 = pam50_colors),
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("white","orange","darkred"))(100),
  main = "Pairwise Pearson Correlation",
  treeheight_row = 30,
  treeheight_col = 30
)
```



```{r}
# Arijaan block
gene_data <- data[, gene_columns] 

# 2. Transpose the matrix (patients become columns, genes become rows)
transposed_genes <- t(gene_data)

# 3. Compute the 1099 x 1099 patient-to-patient correlation matrix
patient_cor_matrix <- cor(transposed_genes, method = "pearson")


corrs <- cor(brca_exp_clean[,-1],method="pearson")
```


### Interpret: does the correlation structure correspond to PAM50 subtypes?

```{r}

pheatmap(corrs, cluster_rows = FALSE,
  cluster_cols = FALSE, show_rownames = F,
  show_colnames = F)

# Jungs ich weiß nicht ob das hier richtig ist, aber so siehts gerade mit dem heatmap aus: -@Arijaan

```


## Step 4 — Hierarchical Clustering
### Cluster samples using hclust.

```{r step4-clustering-and-silhouette, fig.width=10, fig.height=6}

# Load required libraries
library(cluster)
library(ggplot2)
library(tibble) # Required for column_to_rownames
library(dplyr)

# 1. DATA PREPARATION & DISTANCE CALCULATION
# Move gene names to rownames so the matrix is purely numeric
brca_exp_numeric <- brca_exp_clean %>%
  column_to_rownames("gene")  

# Transpose matrix so patients are rows, then calculate Euclidean distance
exp_t <- t(brca_exp_numeric)
D <- dist(exp_t)

# 2. HIERARCHICAL CLUSTERING
hc_result <- hclust(D, method = "ward.D2")

# 3. SILHOUETTE ANALYSIS (Mathematical k)
km <- kmeans(x = exp_t, centers = 2, nstart = 10)
s <- silhouette(km$cluster, D)

cat("Average silhouette for k=2:", mean(s[, 3]), "\n")
plot(s, col = "blue", border = NA, main = "Silhouette Plot for K-means (k=2)")
```

## Cut the dendrogram to obtain k clusters (choose k; justify your choice).

```{r}
# Cut the tree at k = 2 and k = 5
hc_clusters <- cutree(hc_result, k = 2)
hc_clusters_5 <- cutree(hc_result, k = 5) 

plot(hc_result,
     labels = FALSE,
     main = "Hierarchical clustering of BRCA samples",
     xlab = "Samples",
     ylab = "Height")
rect.hclust(hc_result, k = 2, border = "red")

# check cluster sizes for k=2
table(hc_clusters)
```

### Compare cluster membership with PAM50 subtypes.

```{r}
# --- Comparison for k = 2 ---
cluster_df <- data.frame(barcode = names(hc_clusters), 
                         hc_cluster = hc_clusters)

comparison_df <- merge(cluster_df, anot_paper, by = "barcode")

comparison_table <- table(Cluster = comparison_df$hc_cluster, 
                          PAM50 = comparison_df$paper_BRCA_Subtype_PAM50)


# --- Comparison for k = 5 ---
cluster_df_5 <- data.frame(barcode = names(hc_clusters_5), 
                           hc_cluster = hc_clusters_5)

comparison_df_5 <- merge(cluster_df_5, anot_paper, by = "barcode")
comparison_table_5 <- table(Cluster = comparison_df_5$hc_cluster, 
                            PAM50 = comparison_df_5$paper_BRCA_Subtype_PAM50)

print("Cluster Membership vs. PAM50 Subtypes (k=2):")
print(comparison_table)

print("Cluster Membership vs. PAM50 Subtypes (k=5):")
print(comparison_table_5)
```


## Step 5 — PCA Visualization

### Run prcomp on the transposed expression matrix (samples as rows).
```{r}
library(dplyr)
library(tibble)

# The simplest, safest pipeline:
pca_result <- exp_t %>% 
  prcomp(center = TRUE, scale = FALSE)
summary(pca_result)
```



### Plot PC1 vs. PC2 and PC1 vs. PC3, colored by PAM50 subtype and by hierarchical cluster.

```{r}
library(ggplot2)
library(dplyr)

# 1. Create a dataframe and combine it with the annotation
pca_with_annot <- as.data.frame(pca_result$x) %>%
  merge(anot_paper, by.x = 0, by.y = "barcode") 
  # by.x = 0 uses the PCA rownames, by.y = "barcode" uses your annotation column

ggplot(pca_with_annot, aes(x = PC1, y = PC2, color = paper_BRCA_Subtype_PAM50)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  # Optional: You can change the colors manually just like your example
  scale_color_manual(values = c("Basal" = "red", "Her2" = "blue", 
                                "LumA" = "darkgreen", "LumB" = "lightgreen", 
                                "Normal" = "grey"))
ggplot(pca_with_annot, aes(x = PC1, y = PC3, color = paper_BRCA_Subtype_PAM50)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  # Optional: You can change the colors manually just like your example
  scale_color_manual(values = c("Basal" = "red", "Her2" = "blue", 
                                "LumA" = "darkgreen", "LumB" = "lightgreen", 
                                "Normal" = "grey"))

```


### Report the variance explained by the first three PCs. What distinguishes the subtypes?
PC1 captures the most variance in the dataset and mainly distinguishes between Basal and Luminal BRCA tumors. This indicates that the biggest biological difference in breast cancer is between hormone-receptor negative (Basal) and hormone-receptor positive (Luminal) tumors.

Even though the plot shows an overlap between Luminal A and Luminal B tumors, PC2 does seem to distinguish them slightly. Furthermore PC2 also seems to distinguish between HER2- and normal-like subtypes. 
PC3 on the other hand doesnt seem to distinguish between PAM50 subtypes, which indicates that it may distinguish between secondary biological factors, such as Tumor Microenviornment
## Step 6 — Differential Expression

### Use a Wilcoxon rank-sum test (one-vs-rest) to identify marker genes for each PAM50 subtype.

```{r}
wilcox.test()
```


### Apply FDR correction (BH). Report the top 20 genes per subtype by adjusted p-value and log2 fold-change.

### Generate an annotated heatmap (rows = top DEGs, columns = samples, annotation bars = PAM50 subtype and pathologic stage).

## Step 7 — Prediction Models

### Split data 70/30 (stratified by subtype).

```{r}

# Function to remove columns with lots of NAs (more than half of the entries).
# Used later on in the script.
#nacol_rm <- function(df, p){
#  names <- c()
#  valid_colnames <- sapply(1:ncol(df), function(i){
#  ifelse(sum(is.na(df[,i])) > round(nrow(df)*p), NA,
#         names[i]<-colnames(df[,i])
#    )
#   }
#  )

#valid_colnames <- na.omit(valid_colnames)
#df[,valid_colnames]
#}

# ---------alternative to the function: (delete if the function works ---LOL

nacol_rm <- function(df,p){
too_many_nas <- colSums(is.na(df)) > round(nrow(df)*p)
bad_column_names <- colnames(df)[too_many_nas]
df <- df[, !too_many_nas]
}
```

For binary analysis of the data, we need to annotate the subtypes properly, so that we have either a "normal" or a "tumor". For this, we can use `dplyr`:

```{r}

# Coverstion of subtypes into binary:
normal <- brca_anot_clean %>% filter(paper_BRCA_Subtype_PAM50 == "Normal") 
tumor <- brca_anot_clean %>% filter(!paper_BRCA_Subtype_PAM50 == "Normal") 
tumor$paper_BRCA_Subtype_PAM50 <- "tumor"
binary_data <- rbind(normal,tumor)

# Rownames to cols
 
binary_data <- binary_data %>% 
  column_to_rownames("barcode")

binary_data[numeric_cols]
# Convert to numeric
#numeric_cols <- setdiff(names(binary_data), "paper_BRCA_Subtype_PAM5")

#binary_data[numeric_cols] <- lapply(binary_data[,numeric_cols], function(x) as.numeric(as.character(x)))


## AI version:

binary_data <- binary_data %>%
  # Convert everything EXCEPT the target column to numeric
  mutate(across(-paper_BRCA_Subtype_PAM50, ~ as.numeric(as.character(.))))

```


```{r}

# Using caret to create the test and training data
trainIndex <- createDataPartition(binary_data$paper_BRCA_Subtype_PAM50, p = 0.7, list = FALSE)


train_data <- binary_data[trainIndex, ]
test_data  <- binary_data[-trainIndex, ]

# Clearing out columns with lots of NAs 
train_data <- nacol_rm(df=train_data, p=0.5) # proportion of cols with NAs
test_data <- nacol_rm(df=test_data, p=0.5)

```


```{r}

##################### Gemini glm #############################

# 1. Conversion of subtypes into binary
normal <- brca_anot_clean %>% filter(paper_BRCA_Subtype_PAM50 == "Normal") 
tumor <- brca_anot_clean %>% filter(paper_BRCA_Subtype_PAM50 != "Normal") # Safer than !col == "Normal"
tumor$paper_BRCA_Subtype_PAM50 <- "tumor"
binary_data <- rbind(normal, tumor)

# 2. Rownames to cols
binary_data <- binary_data %>% 
  column_to_rownames("barcode")

# 3. Handle NAs and Factors safely at the dataframe level
binary_data <- binary_data %>% 
  filter(!is.na(paper_BRCA_Subtype_PAM50)) %>% 
  mutate(paper_BRCA_Subtype_PAM50 = factor(paper_BRCA_Subtype_PAM50, levels = c("Normal", "tumor")))

# 4. Get numeric columns (Typo fixed)
numeric_cols <- setdiff(names(binary_data), "paper_BRCA_Subtype_PAM50")

# (Optional: Convert to numeric if you uncommented your lapply)
# binary_data[numeric_cols] <- lapply(binary_data[numeric_cols], function(x) as.numeric(as.character(x)))

# 5. Split Data
set.seed(123) # Always set a seed for reproducibility
trainIndex <- createDataPartition(binary_data$paper_BRCA_Subtype_PAM50, p = 0.7, list = FALSE)

train_data <- binary_data[trainIndex, ]
test_data  <- binary_data[-trainIndex, ]

# 6. Correct NA handling (Preventing Data Leakage)
# Calculate which columns to drop based ONLY on training data
too_many_nas <- colSums(is.na(train_data)) > round(nrow(train_data) * 0.5)
cols_to_drop <- colnames(train_data)[too_many_nas]

# Drop the exact same columns from BOTH datasets
train_data <- train_data[, !colnames(train_data) %in% cols_to_drop]
test_data  <- test_data[, !colnames(test_data) %in% cols_to_drop]

# 7. Model
model_logit <- glm(paper_BRCA_Subtype_PAM50 ~ ., data = train_data, family = "binomial")

summary(model_logit)
```

### Binary task — tumor vs. normal: Logistic Regression, Random Forest, k-NN. Report confusion matrices and AUC-ROC.

#### Logistic regression

```{r Logistic regression}

# Creating the model using the glm function from the stats library.

# NAs can be inputted here since the function takes out all NAs. 
model_logit <- glm(paper_BRCA_Subtype_PAM50 ~ ., data = train_data, family = "binomial")

summary(model_logit)

# Predict
prob_logit <- predict(model_logit, test, type = "response")
pred_logit <- ifelse(prob_logit > 0.5, "normal", "tumor")
pred_logit <- factor(pred_logit, levels = c("normal", "tumor"))

confusionMatrix(pred_logit$paper_vital_status, test_data$paper_vital_status)

```

#### K-Nearest Neighbor:
```{r}
# Train k-NN
set.seed(123)
model_knn <- train(
  Class ~ .,
  data = train_norm,
  method = "knn",
  tuneGrid = expand.grid(k = 1:10),
  trControl = trainControl(method = "cv", number = 5, classProbs = TRUE)
)
```

```{r}
# Predict
knn_probs <- predict(model_knn, test_norm, type = "prob")[, "malignant"]
pred_knn <- ifelse(knn_probs > 0.5, "malignant", "benign")
pred_knn <- factor(pred_knn, levels = c("malignant", "benign"))
confusionMatrix(pred_knn, test_norm$Class)

```
```{r}
# Normalize features (excluding Class)
preproc <- preProcess(train_data[, -10], method = c("center", "scale"))
train_norm <- predict(preproc, train_data[, -10])
test_norm <- predict(preproc, test_data[, -10])

# Add Subtype back
train_norm <- train_data
test_norm <- test_data
```


```{r}
# Plot k selection
ggplot(model_knn$results, aes(x = k, y = Accuracy)) +
  geom_line(col="red3") +
  geom_point() +
  labs(title = "k-NN Accuracy vs. Number of Neighbors")
```

#### Random forest


-----------------------------------------------------------------



### Multiclass task — predict PAM50 subtype: Random Forest, Multinomial Logistic Regression, k-NN.

#### Random Forest:

```{r}
library(randomForest)
# appending meaningfull expression data of the top 100 genes to the annotations dataset for increasing accuarcy.

sig_genes <- head(gene_variability, 100)
sig_genes <- column_to_rownames(sig_genes, "gene")
dat <- brca_anot_clean %>% 
  column_to_rownames("barcode") %>% 
  filter(!is.na(paper_BRCA_Subtype_PAM50)) %>% 
  mutate(exp = t(sig_genes))
# Fit the random forest model
# Using caret to create the test and training data

trainIndex <- createDataPartition(dat$paper_BRCA_Subtype_PAM50, p = 0.7, list = FALSE)


train_data <- binary_data[trainIndex, ]
test_data  <- binary_data[-trainIndex, ]

# Clearing out columns with lots of NAs 
train_data <- nacol_rm(df=train_data, p=0.5) # proportion of cols with NAs
test_data <- nacol_rm(df=test_data, p=0.5)

model_rf <- randomForest(paper_BRCA_Subtype_PAM50 ~ ., data = train_data, importance = TRUE, ntree = 500)

# Feature importance plot
varImpPlot(model_rf, main = "Feature Importance")

# Predict
pred_rf <- predict(model_rf, test_data)
confusionMatrix(pred_rf, test_data$paper_br) # enter subtype 

confusionMatrix(pred_rf, test_data) # enter subtype
```

### For each model, report: accuracy, F1, per-class precision/recall.

## Step 8 — Method Comparison

### Compare model performance systematically in a summary table.

### Evaluate whether the top DEGs (Step 6) rank highly as predictive features (Step 7 feature importance). Overlap analysis with UpSet plot or Venn diagram.

# Part II — Clinical Data Analysis

## Step A — Annotation Cleaning

### Tidy the annotation file to keep only curated, paper-level columns.

### anot_clean \<- brca_anot_clean[, c("barcode", grep("\^paper\_", names(brca_anot_clean), value = TRUE))]

### How many paper\_ columns are retained? What categories of information do they cover?
```{r}
# Count the number of paper_ columns
num_paper_cols <- ncol(anot_paper) - 1
print(num_paper_cols)

# View the column names to see the categories
colnames(anot_paper)
```

### Are there samples with missing subtype assignments? How should these be handled?
```{r}
# Check for missing values in the subtype column
table(anot_paper$paper_BRCA_Subtype_PAM50, useNA = "ifany")

# Or count NAs directly
sum(is.na(anot_paper$paper_BRCA_Subtype_PAM50))
```


## Step B — Clinical Variable Exploration

### Explore the clinical metadata as a standalone dataset before merging with expression data.

### Explore paper_pathologic_stage, paper_BRCA_Pathology (IDC vs ILC), paper_age_at_initial_pathologic_diagnosis.

```{r}
## Patholgical Stage
plot_stage <- anot_paper %>%
  drop_na(paper_pathologic_stage) %>%
  ggplot(aes(x = paper_pathologic_stage)) +
  geom_bar(fill = "red", alpha = 0.7) +
  labs(title = "Pathologic Stage", x = "Stage", y = "Count")
plot_age <- ggplot(anot_paper, aes(x = paper_age_at_initial_pathologic_diagnosis)) +
  geom_histogram(fill = "steelblue", color = "black", bins = 25, alpha = 0.7) +
  labs(title = "Age at Diagnosis", x = "Age (Years)", y = "Count")
print(plot_age)
print(plot_stage)
cat("--- BRCA Pathology (IDC vs ILC) ---\n")
print(table(anot_paper$paper_BRCA_Pathology, useNA = "ifany"))

```


### Tabulate paper_vital_status and compute median follow-up time.
```{r}
cat("--- Vital Status ---")
print(table(anot_paper$paper_vital_status, useNA = "ifany"))

median_followup_days <- median(anot_paper$paper_days_to_last_followup, na.rm = TRUE)
print(median_followup_days)
```
### Examine multi-omic cluster columns (paper_CNV.Clusters, paper_mRNA.Clusters, paper_Protein.Clusters) — how do they relate to PAM50?
```{r}
cat("--- mRNA Clusters vs PAM50 ---\n")
print(table(Cluster = anot_paper$paper_mRNA.Clusters, PAM50 = anot_paper$paper_BRCA_Subtype_PAM50))
cat("\n")

cat("--- CNV (Copy Number) Clusters vs PAM50 ---\n")
print(table(Cluster = anot_paper$paper_CNV.Clusters, PAM50 = anot_paper$paper_BRCA_Subtype_PAM50))
cat("\n")

cat("--- Protein Clusters vs PAM50 ---\n")
print(table(Cluster = anot_paper$paper_Protein.Clusters, PAM50 = anot_paper$paper_BRCA_Subtype_PAM50))
cat("\n")
```
 
## Step C — Clinical Feature Classification

### Predict a binary or multi-class clinical outcome using clinical variables only (no expression data).

### Define the outcome variable and apply appropriate encoding.

We define vital status as outcome variable (predicting whether a patient is 'Alive' or 'Dead' based on their clinical profile)

### Select a small set of meaningful features (age, stage, grade, subtype, etc.) and handle missing values by imputation or exclusion.

```{r}
# We select features, drop missing values, and encode categorical data as factors
ml_data <- anot_paper %>%
  select(paper_vital_status, paper_age_at_initial_pathologic_diagnosis, 
         paper_pathologic_stage, paper_BRCA_Subtype_PAM50) %>%
  drop_na() %>%
  mutate(
    Target = factor(paper_vital_status, levels = c("Alive", "Dead")),
    Age = paper_age_at_initial_pathologic_diagnosis,
    Stage = as.factor(paper_pathologic_stage),
    PAM50 = as.factor(paper_BRCA_Subtype_PAM50)
  ) %>%
  select(-paper_vital_status, -paper_age_at_initial_pathologic_diagnosis, 
         -paper_pathologic_stage, -paper_BRCA_Subtype_PAM50)

```

### Train/test split (70/30 or 5-fold CV). Fit: Logistic Regression, Random Forest, k-NN.

```{r}
## TRAIN / TEST SPLIT
set.seed(42) # For reproducible splits

trainIndex <- createDataPartition(ml_data$Target, p = 0.7, list = FALSE)
train_data <- ml_data[trainIndex, ]
test_data  <- ml_data[-trainIndex, ]

## MODEL TRAINING (with 5-fold CV)

cv_ctrl <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary, savePredictions = "final")

lr_model <- train(Target ~ ., data = train_data, method = "glm", family = "binomial",
                  trControl = cv_ctrl, metric = "ROC")
rf_model <- train(Target ~ ., data = train_data, method = "rf", 
                  trControl = cv_ctrl, metric = "ROC")
# k-NN relies on distance, so we MUST center and scale the Age variable automatically
knn_model <- train(Target ~ ., data = train_data, method = "knn", 
                   trControl = cv_ctrl, metric = "ROC", preProcess = c("center", "scale"))

print(lr_model)
print(rf_model)
print(knn_model)
```


### Report confusion matrices, accuracy, sensitivity/specificity, and AUC-ROC for each model.
```{r}

set.seed(42) 
## PREDICTIONS & CONFUSION MATRICES

# Logistic Regression
pred_lr_class <- predict(lr_model, newdata = test_data)
pred_rf_class <- predict(rf_model, newdata = test_data)
pred_knn_class <- predict(knn_model, newdata = test_data)

# Print Confusion Matrices (Includes Accuracy, Sensitivity, Specificity)
cat("\n--- Logistic Regression Confusion Matrix ---\n")
confusionMatrix(pred_lr_class, test_data$Target)

cat("\n--- Random Forest Confusion Matrix ---\n")
confusionMatrix(pred_rf_class, test_data$Target)

cat("\n--- k-NN Confusion Matrix ---\n")
confusionMatrix(pred_knn_class, test_data$Target)
              
```
```{r}

set.seed(42) 
# AUC-ROC AND PLOTTING

# Get predicted probabilities for the test set (using "Alive" as the event of interest)
prob_lr  <- predict(lr_model, newdata = test_data, type = "prob")
prob_rf  <- predict(rf_model, newdata = test_data, type = "prob")
prob_knn <- predict(knn_model, newdata = test_data, type = "prob")

# Compute ROC
roc_logit <- roc(response = test_data$Target, predictor = prob_lr$Alive, levels = c("Alive", "Dead"))
roc_rf    <- roc(response = test_data$Target, predictor = prob_rf$Alive, levels = c("Alive", "Dead"))
roc_knn   <- roc(response = test_data$Target, predictor = prob_knn$Alive, levels = c("Alive", "Dead"))

# Print AUC values
cat("\n--- AUC Values ---\n")
cat("Logistic Regression AUC:", roc_logit$auc, "\n")
cat("Random Forest AUC:", roc_rf$auc, "\n")
cat("k-NN AUC:", roc_knn$auc, "\n")

# Plot ROC curves
plot(roc_logit, col = "blue", main = "ROC Curves (Test Set)", lwd = 2)
lines(roc_rf, col = "green", lwd = 2)
lines(roc_knn, col = "red", lwd = 2)
legend("bottomright", legend = c("Logistic Regression", "Random Forest", "k-NN"),
       col = c("blue", "green", "red"), lwd = 2)
```



### Identify the most predictive clinical features: variable importance from RF, and coefficients from Logistic Regression. Note that k-NN does not provide native feature importance.
```{r}
set.seed(42) 
# Random Forest Variable Importance Plot
rf_importance <- varImp(rf_model, scale = FALSE)
plot(rf_importance, main = "Random Forest Feature Importance")

# Logistic Regression Coefficients / Importance
# Using summary() to see raw coefficients and p-values
cat("\n--- Logistic Regression Coefficients ---\n")
summary(lr_model$finalModel)

# Optional: Plot Logistic Regression Variable Importance
lr_importance <- varImp(lr_model, scale = FALSE)
plot(lr_importance, main = "Logistic Regression Feature Importance")
```


## Step D — Integrative Analysis (Expression + Clinical)

### Correlate top DEGs from Part I with continuous clinical variables (e.g., age) using Spearman's rho.

### Test whether top DEGs differ significantly across clinical stages.

### Build a multimodal classifier: combine top expression features (from Step 7 / RF importance) with key clinical features. Does adding clinical data improve performance?

### Visualize: boxplots of top expression features stratified by clinical subgroups; annotated heatmap with both molecular and clinical annotation bars.

### 

## Step E - Discussion & Interpretation

### Critical thinking questions

### Are the molecular subtypes (from expression clustering) concordant with clinical staging? Where do they disagree?

### Which data modality — expression or clinical — gives better predictive accuracy for your outcome variable?

### What biological mechanisms might explain the top DEGs found between clinical subgroups?

### What are the limitations of the TCGA cohort for clinical prediction (sample size per stage, selection of academic-centre patients, etc.)?
