---
title: "BRCA-Project"
output: html_notebook
---

# PROJECT 1 — TCGA-BRCA \| Breast Invasive Carcinoma

## Cleaning annotations

```{r}
load("BRCA_data.RData")
library(dplyr) 
library(ggplot2) 
library(pheatmap) 
library(caret)
library (tibble)
library(tidyr)
```

# Part I - Gene Expression Analysis


## How many samples and genes are in the expression matrix?

```{r}
# The dim() function returns the dimensions of your dataset: [1] Number of Rows, [2] Number of Columns.

dim <- dim(brca_exp_clean)
df_dims <- data.frame("Samples" = dim[1], "Genes" = dim[2])

## Alternative
#num_genes <- nrow(brca_exp_clean)
#num_samples_exp <- ncol(brca_exp_clean)
#cat("Expression matrix has", num_genes, "genes and", num_samples_exp, "samples.\n")
#cat("Retained paper_ columns:", ncol(anot_paper) - 1, "\n")

```

## significant breast cancer genes:

How did we get these genes? @kareembeltagi7-droid

PIK3CA, PTEN, AKT1, TP53, GATA3, CDH1, RB1, MLL3, MAP3K1, CDKN1B, TBX3, RUNX1, CBFB, AFF2, PIK3R1, PTPN22, PTPRD, NF1, SF3B1 and CCND3

## How many annotation columns are retained after cleaning to paper\_ columns?

```{r}
# indexes the annotations for the columns starting with paper, adding a barcode column for counting.
anot_paper <- brca_anot_clean[, c("barcode", grep("^paper_", names(brca_anot_clean), value = TRUE))]

# check how many columns start with paper_
num_paper_cols <- ncol(anot_paper) - 1 # Subtract 1 to exclude the 'barcode' column
cat("Number of paper_ columns retained:", num_paper_cols, "\n")
```

## What is the sample count per PAM50 subtype (paper_BRCA_Subtype_PAM50)?

```{r}
# Shows all PAM50 subtypes in a table. The "ifany" argument adds a column with the NA count if there are any NAs available.

subtype_counts <- table(anot_paper$paper_BRCA_Subtype_PAM50, useNA = "ifany")

print("Sample count per PAM50 subtype:")
print(subtype_counts)

## Finding the Valid Patients
# The primary target for BRCA is paper_BRCA_Subtype_PAM50
valid_target_samples <- sum(!is.na(anot_paper$paper_BRCA_Subtype_PAM50))
cat("Samples with non-NA PAM50 subtype:", valid_target_samples, "\n")

```

=======

# Part I - Gene Expression Analysis 

## How many samples and genes are in the expression matrix?

```{r}
# --- 1. Samples and Genes in the Expression Matrix ---
dim(brca_exp_clean)
num_genes <- nrow(brca_exp_clean)
num_samples_exp <- ncol(brca_exp_clean)
cat("Expression matrix has", num_genes, "genes and", num_samples_exp, "samples.\n")
cat("Retained paper_ columns:", ncol(anot_paper) - 1, "\n")
```


## significant breast cancer genes:

PIK3CA, PTEN, AKT1, TP53, GATA3, CDH1, RB1, MLL3, MAP3K1, CDKN1B, TBX3, RUNX1, CBFB, AFF2, PIK3R1, PTPN22, PTPRD, NF1, SF3B1 and CCND3


## How many annotation columns are retained after cleaning to paper\_ columns?

```{r}
# check how many columns start with paper_
num_paper_cols <- ncol(anot_paper) - 1 # Subtract 1 to exclude the 'barcode' column
cat("Number of paper_ columns retained:", num_paper_cols, "\n")
```


## What is the sample count per PAM50 subtype (paper_BRCA_Subtype_PAM50)?

```{r}
## Sample Count per PAM50 Subtype

subtype_counts <- table(anot_paper$paper_BRCA_Subtype_PAM50, useNA = "ifany")

print("Sample count per PAM50 subtype:")
print(subtype_counts)
```
 
## Step 2 — Expression Distribution Analysis

### Randomly select 10 genes and visualize their expression distributions across all samples.

```{r}

set.seed(161) # For reproducibility
random_exp <- brca_exp_clean %>% slice_sample(n=10) # Subsetting the matrix to keep randomly selected genes
random_exp_long <- random_exp %>%
  pivot_longer(cols = where(is.numeric), 
               names_to = "Sample", 
               values_to = "Expression") # Reshaping data for ggplot2

ggplot(random_exp_long, aes(x = gene, y = Expression, fill = gene)) +
geom_boxplot(alpha = 0.7, outlier.size = 0.5) +  
  theme_minimal() +                                
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"), 
        legend.position = "none") +                
  labs(title = "Expression Distribution of 10 Randomly Selected Genes",
       x = "Gene",
       y = "Normalized Expression Level")


```


### Select the top 5 and bottom 5 most variable genes (by standard deviation); plot their distributions and compare.
```{r}
# Calculate gene variability (standard deviation)
# Calculate standard deviation for each gene (across all samples)
# Exclude the "gene" column, only use expression columns
gene_variability <- brca_exp_clean %>%
  mutate(gene_sd = apply(select(., -gene), 1, sd, na.rm = TRUE)) %>%
  select(gene, gene_sd) %>%
  arrange(desc(gene_sd))

# Top 5 most variable
top5_variable <- head(gene_variability, 5)
cat("Top 5 Most Variable Genes:\n")
print(top5_variable)

# Bottom 5 least variable
bottom5_variable <- tail(gene_variability, 5)
cat("\nBottom 5 Least Variable Genes:\n")
print(bottom5_variable)

# Combine gene lists
variable_genes <- c(top5_variable$gene, bottom5_variable$gene)

# Extract and transform data
var_exp <- brca_exp_clean %>%
  filter(gene %in% variable_genes) %>%
  pivot_longer(-gene, names_to = "Sample", values_to = "Expression") %>%
  mutate(Variability = ifelse(gene %in% top5_variable$gene, "High", "Low"))

# Plot comparison
ggplot(var_exp, aes(x = reorder(gene, Expression, FUN = median), 
                    y = Expression, 
                    fill = Variability)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_manual(values = c("High" = "#E74C3C", "Low" = "#3498DB")) +
  labs(title = "Expression Distribution: High vs Low Variability Genes",
       subtitle = "Top 5 and Bottom 5 genes by standard deviation",
       x = "Gene",
       y = "Normalized Expression (log2)",
       fill = "Variability") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## Step 3 — Correlation Analysis

### Compute pairwise Pearson correlations across all samples using the full gene set.

### Visualize as a correlation heatmap. What structure do you observe? Are there sub- blocks?

### Interpret: does the correlation structure correspond to PAM50 subtypes?

## Step 4 — Hierarchical Clustering

### Cluster samples using hclust.

### Cut the dendrogram to obtain k clusters (choose k; justify your choice).

### Compare cluster membership with PAM50 subtypes.

## Step 5 — PCA Visualization

### Run prcomp on the transposed expression matrix (samples as rows).

### Plot PC1 vs. PC2 and PC1 vs. PC3, colored by PAM50 subtype and by hierarchical cluster.

### Report the variance explained by the first three PCs. What distinguishes the subtypes?

## Step 6 — Differential Expression

### Use a Wilcoxon rank-sum test (one-vs-rest) to identify marker genes for each PAM50 subtype.

### Apply FDR correction (BH). Report the top 20 genes per subtype by adjusted p-value and log2 fold-change.

### Generate an annotated heatmap (rows = top DEGs, columns = samples, annotation bars = PAM50 subtype and pathologic stage).

## Step 7 — Prediction Models

### Split data 70/30 (stratified by subtype).

```{r}

# Function to remove columns with lots of NAs (more than half of the entries).
# Used later on in the script.
nacol_rm <- function(df, p){
  names <- c()
  valid_colnames <- sapply(1:ncol(df), function(i){
  ifelse(sum(is.na(df[,i])) > round(nrow(df)*p), NA,
         names[i]<-colnames(df[,i])
    )
   }
  )
valid_colnames <- na.omit(valid_colnames)
df[,valid_colnames]
}

# ---------alternative to the function: (delete if the function works lol)-----
#too_many_nas <- colSums(is.na(train_data)) > 500
#bad_column_names <- colnames(train_data)[too_many_nas]
#train_data_filtered <- train_data[, !too_many_nas]

```

```{r}

# Using caret to create the test and training data
trainIndex <- createDataPartition(brca_anot_clean$paper_BRCA_Subtype_PAM50, p = 0.7, list = FALSE)


train_data <- brca_anot_clean[trainIndex, ]
test_data  <- brca_anot_clean[-trainIndex, ]

# Clearing out columns with lots of NAs 
train_data <- nacol_rm(df=train_data, p=0.5)
test_data <- nacol_rm(df=test_data, p=0.5)


```

### Binary task — tumor vs. normal: Logistic Regression, Random Forest, k-NN. Report confusion matrices and AUC-ROC.

The provided dataset only has breast cancer patients and there are no healthy controls available to compare the patients to. Instead, we compare another binary variable: vital status. Here the patient can only be "Alive" or "Dead":

```{r}

vitals <- unique(brca_anot_clean$paper_vital_status) ; vitals() 



```


#### Logistic regression

```{r}
# Creating the model using the glm function from the stats library.
## TODO for @Arijaan: replace Class with SUBTYPE.

model_logit <- glm(paper_vital_status ~ ., data = train_data, family = "binomial")
summary(model_logit)

# Predict
prob_logit <- predict(model_logit, test, type = "response")
pred_logit <- ifelse(prob_logit > 0.5, "alive", "dead")
pred_logit <- factor(pred_logit, levels = c("alive", "dead"))

confusionMatrix(pred_logit$paper_vital_status, test_data$paper_vital_status)

```

#### K-Nearest Neighbor:

```{r}
# Normalize features (excluding Class)
preproc <- preProcess(train_data[, -10], method = c("center", "scale"))
train_norm <- predict(preproc, train_data[, -10])
test_norm <- predict(preproc, test_data[, -10])

# Add Subtype back
train_norm <- train_data
test_norm <- test_data
```

### Multiclass task — predict PAM50 subtype: Random Forest, Multinomial Logistic Regression, k-NN.

#### Random Forest:

```{r}
# Fit the random forest model
model_rf <- randomForest(Class ~ ., data = train_data, importance = TRUE, ntree = 5000)

# Feature importance plot
varImpPlot(model_rf, main = "Feature Importance")

# Predict
pred_rf <- predict(model_rf, test_data)
confusionMatrix(pred_rf, test_data) # enter subtype 

confusionMatrix(pred_rf, test_data) # enter subtype
```

```{r}
# Train k-NN
set.seed(123)
model_knn <- train(
  Class ~ .,
  data = train_norm,
  method = "knn",
  tuneGrid = expand.grid(k = 1:10),
  trControl = trainControl(method = "cv", number = 5, classProbs = TRUE)
)
```

```{r}
# Predict
knn_probs <- predict(model_knn, test_norm, type = "prob")[, "malignant"]
pred_knn <- ifelse(knn_probs > 0.5, "malignant", "benign")
pred_knn <- factor(pred_knn, levels = c("malignant", "benign"))
confusionMatrix(pred_knn, test_norm$Class)

```

```{r}
# Plot k selection
ggplot(model_knn$results, aes(x = k, y = Accuracy)) +
  geom_line(col="red3") +
  geom_point() +
  labs(title = "k-NN Accuracy vs. Number of Neighbors")
```

### For each model, report: accuracy, F1, per-class precision/recall.

## Step 8 — Method Comparison

### Compare model performance systematically in a summary table.

### Evaluate whether the top DEGs (Step 6) rank highly as predictive features (Step 7 feature importance). Overlap analysis with UpSet plot or Venn diagram.

# Part II — Clinical Data Analysis

## Step A — Annotation Cleaning

### Tidy the annotation file to keep only curated, paper-level columns.

### anot_clean \<- brca_anot_clean[, c("barcode", grep("\^paper\_", names(brca_anot_clean), value = TRUE))]

### How many paper\_ columns are retained? What categories of information do they cover?

### Are there samples with missing subtype assignments? How should these be handled?

## Step B — Clinical Variable Exploration

### Explore the clinical metadata as a standalone dataset before merging with expression data.

### Explore paper_pathologic_stage, paper_BRCA_Pathology (IDC vs ILC), paper_age_at_initial_pathologic_diagnosis.

### Tabulate paper_vital_status and compute median follow-up time.

### Examine multi-omic cluster columns (paper_CNV.Clusters, paper_mRNA.Clusters, paper_Protein.Clusters) — how do they relate to PAM50?

## Step C — Clinical Feature Classification

### Predict a binary or multi-class clinical outcome using clinical variables only (no expression data).

### Define the outcome variable and apply appropriate encoding.

### Select a small set of meaningful features (age, stage, grade, subtype, etc.) and handle missing values by imputation or exclusion.

### Train/test split (70/30 or 5-fold CV). Fit: Logistic Regression, Random Forest, k-NN.

### Report confusion matrices, accuracy, sensitivity/specificity, and AUC-ROC for each model.

### Identify the most predictive clinical features: variable importance from RF, and coefficients from Logistic Regression. Note that k-NN does not provide native feature importance.

## Step D — Integrative Analysis (Expression + Clinical)

### Correlate top DEGs from Part I with continuous clinical variables (e.g., age) using Spearman's rho.

### Test whether top DEGs differ significantly across clinical stages.

### Build a multimodal classifier: combine top expression features (from Step 7 / RF importance) with key clinical features. Does adding clinical data improve performance?

### Visualize: boxplots of top expression features stratified by clinical subgroups; annotated heatmap with both molecular and clinical annotation bars.

### 

## Step E - Discussion & Interpretation

### Critical thinking questions

### Are the molecular subtypes (from expression clustering) concordant with clinical

staging? Where do they disagree?

### Which data modality — expression or clinical — gives better predictive accuracy for

your outcome variable?

### What biological mechanisms might explain the top DEGs found between clinical

subgroups?

### What are the limitations of the TCGA cohort for clinical prediction (sample size per

stage, selection of academic-centre patients, etc.)?
