---
title: "BRCA-Project"
output: html_notebook
---

# PROJECT 1 — TCGA-BRCA \| Breast Invasive Carcinoma

## Cleaning annotations

```{r}
load("~/BRCA-Project/BRCA_data.RData")
library(dplyr) 
library(ggplot2) 
library(pheatmap) 
library(caret) 
```

```{r}

#-------------------------Cleaning annotations---------------------------------
anot_paper <- brca_anot_clean[, c("barcode", grep("^paper_", names(brca_anot_clean), value = TRUE))]

# check how many columns start with paper_
num_paper_cols <- ncol(anot_paper) - 1 # Subtract 1 to exclude the 'barcode' column
cat("Number of paper_ columns retained:", num_paper_cols, "\n")

## Finding the Valid Patients
# The primary target for BRCA is paper_BRCA_Subtype_PAM50
valid_target_samples <- sum(!is.na(anot_paper$paper_BRCA_Subtype_PAM50))
cat("Samples with non-NA PAM50 subtype:", valid_target_samples, "\n")

```

## Part 1: Gene Expression Analysis

### Data structure investigation:

```{r step1-data-structure}
# --- 1. Samples and Genes in the Expression Matrix ---
# The dim() function returns the dimensions of your dataset: [1] Number of Rows, [2] Number of Columns.
dim(brca_exp_clean)

num_genes <- nrow(brca_exp_clean)
num_samples_exp <- ncol(brca_exp_clean)
cat("Expression matrix has", num_genes, "genes and", num_samples_exp, "samples.\n")
cat("Retained paper_ columns:", ncol(anot_paper) - 1, "\n")

## 3. Sample Count per PAM50 Subtype

subtype_counts <- table(anot_paper$paper_BRCA_Subtype_PAM50, useNA = "ifany")

print("Sample count per PAM50 subtype:")
print(subtype_counts)
```

## How many samples and genes are in the expression matrix?

## significant breast cancer genes:

PIK3CA, PTEN, AKT1, TP53, GATA3, CDH1, RB1, MLL3, MAP3K1, CDKN1B, TBX3, RUNX1, CBFB, AFF2, PIK3R1, PTPN22, PTPRD, NF1, SF3B1 and CCND3

## Step E - Discussion & Interpretation

### Critical thinking questions

### Are the molecular subtypes (from expression clustering) concordant with clinical

staging? Where do they disagree?

### Which data modality — expression or clinical — gives better predictive accuracy for

your outcome variable?

### What biological mechanisms might explain the top DEGs found between clinical

subgroups?

### What are the limitations of the TCGA cohort for clinical prediction (sample size per

stage, selection of academic-centre patients, etc.)?
